### 2016-01-20 02-related.tex
updated notations in section 2

% In this section, we briefly review the Bayesian model we are considering in this paper.
\subsection{Assortative Mixed-membership Stochastic Blockmodels (a-MMSB)}
The assortative mixed membership stochastic blockmodel (a-MMSB)~\cite{gopalan2012scalable} is a special case of MMSB~\cite{airoldi2009mixed} that models the group structure in a network of $N$ vertices. Consider a set  $\cV^*$ containing all the vertices in the graph, and a set $\cE^*$ containing all the linked edges between pairs of vertices. Each vertex $a$ in the vertex set $\cV^*$ has a $K$-dimensional probability distribution $\pi_a$ of participating in the $K$ members of the community set $\cK$.  For every possible peer $b$ in the network, each vertex $a$ randomly draws a community $z_{ab}$. If a pair of vertices ($a,b$) in the edge set $\cE^*$ are in the same community, i.e., $z_{ab}=z_{ba} = k$, then they have a significant probability $\bt_k$ to connect, i.e., $y_{ab}=1$. Otherwise this probability is small. Each community has its connection strength $\beta_{k} \in (0,1)$ which reflects how likely its members are linked to each other. The whole generative process of a-MMSB is then described by:
\benum
\item For each community $k$, draw community strength $\bt_k \sim \mbox{Beta}(\eta)$
\item For each vertex $a$, draw community memberships $\pi_a \sim \mbox{Dirichlet}(\al)$
\item For each pair of vertices $a$ and $b$,
\benum
	\item Draw interaction indicator $z_{ab} \sim \pi_a$
    \item Draw interaction indicator $z_{ba} \sim \pi_b$
    \item Draw link $y_{ab} \sim \mbox{Bernoulli}(r)$, where $r = \bt_k$ if $z_{ab}=z_{ba}=k$, and $r=\dt$ otherwise.
\eenum
\eenum
$\eta$ and $\alpha$ are parameters to the Beta and Dirichlet distribution functions. $\dt$ (usually small) is a parameter for the algorithm.

\newcommand{\Vertices}{\ensuremath{\cV^*{}}\xspace}
\newcommand{\Edges}{\ensuremath{\cE^*{}}\xspace}
\newcommand{\Heldout}{\ensuremath{\cE_h{}}\xspace}
\newcommand{\SGLDMinibatch}{\ensuremath{\cD_n{}}\xspace}
\newcommand{\Minibatch}{\ensuremath{\cE_n{}}\xspace}
\newcommand{\Minibatcht}{\ensuremath{\cE_{n_t}{}}\xspace}
\newcommand{\Neighbors}{\ensuremath{\cV_n{}}\xspace}

\subsection{Stochastic Gradient Markov Chain Monte Carlo}
The algorithm we are working on in this paper is based on stochastic gradient Langevin dynamics (SGLD)~\cite{welling2011bayesian}. SGLD applies the following update rule to obtain samples from a posterior distribution $p(\ta|\cX) \propto p(\cX|\ta) p(\ta)$ of $N$ i.i.d. data points $\cX=\{x_i\}_{i=1}^{N}$:
\bea
\ta^* \law \ta + \f{\epsilon_t}{2}\left(\nabla_{\ta}\log p(\ta_t)+N\bar{g}(\ta;\SGLDMinibatch)\right) + \xi_t, \label{eqn:sgld_update}
\eea
where $\xi_t \sim \cN(0,\epsilon_t )$ with $\ep_t$ the step size, $\SGLDMinibatch$ a mini-batch of size $n$ sampled from $\cX$, and $\bar{g}(\ta;\SGLDMinibatch)$ the mean stochastic gradient, i.e., $\frac{1}{|\SGLDMinibatch|}\sum_{x\in \SGLDMinibatch}^{}\grad_{\ta} \log p(x|\ta)$. As the step size goes to zero by a schedule satisfying $\sm{t}{\infty}\ep_t = \infty$ and $\sm{t}{\infty}\ep_t^2 < \infty$, SGLD samples from the true posterior distribution. One benefit of using SGLD is that we do not need the Metropolis-Hastings (MH) accept-reject tests since the rejection probability goes to zero as the step size collapses to zero. Although the finite step size, which we use in practice rather than converging to zero, still results in some bias, we can reduce the overall error by drawing many more samples per unit time.

SGLD originated from the Langevin Monte Carlo (LMC)~\cite{girolami2011riemann}, where unlike SGLD the gradient is computed exactly by using all data points. Then, Metropolis-Hastings accept-reject tests are applied. Comparing to LMC, SGLD only requires to process a mini-batch $\SGLDMinibatch$ at each iteration and ignores the MH test, and thus the computational complexity substantially reduces from $\cO(N)$ to $\cO(n)$.

@@

\paragraph{\textbf{Sampling global parameters}} The update rule for global parameter $\theta$ is
\bea
\ta_{ki}^* \law \left| \ta_{ki} + \f{\ep}{2} \left( \eta - \ta_{ki} + h(\Minibatcht)\sum_{(a,b) \in \Minibatcht}g_{ab}(\ta_{ki})\right) \right.\nn \\ \left.+ (\ta_{ki})^{\ha}\xi_{ki} \right|, \label{eqn:global_update}
\eea
where the gradient in $\theta$
\bea
g_{ab}(\ta_{ki})
&=& \f{f_{ab}^{(y)}(k,k)}{Z_{ab}^{(y)}} \left(\f{|1-i-y|}{\ta_{ki}} - \f{1}{\sum_j\ta_{kj}} \right),
\eea
here $\Minibatcht$ is a mini-batch of $n_t$ vertex pairs sampled from $\Edges$ or $E$; $h(\Minibatcht)$ is a weight factor to scale the effect of a mini-batch towards the full network; and 
\bea
f_{ab}^{(y)}(k,l)=
\begin{cases}
\bt_k^y(1-\bt_k)^{(1-y)}\pi_{ak}\pi_{bk}, & \mbox{if } k=l\\
\dt^{y}(1-\dt)^{(1-y)} \pi_{ak} \pi_{bl}, & \mbox{if } k \neq l \nn.
\end{cases}
\label{eqn:case}
\eea
$Z_{ab}^{(y)}$ is the normalization constant which we can compute in $\cO(K)$ time \cite{LiAW15}.

\paragraph{\textbf{Sampling local parameters}}
The update rule for the local parameters $\phi$ is
\bea
\phi_{ak}^* \law \left| \phi_{ak} + \f{\ep}{2} \left( \al - \phi_{ak} + \f{N}{|\Neighbors|} \sum_{b \in \Neighbors} g_{ab}(\phi_{ak})\right) \right. \nn \\ \left. + (\phi_{ak})^{\ha} \xi_{ak}\right|,
\label{eqn:local_update}
\eea
where the gradient in $\phi$
\bea
g_{ab}(\phi_{ak}) = \f{f_{ab}^{(y)}(k)}{Z_{ab}^{(y)}\phi_{ak}} - \f{1}{\sum_j\phi_{aj}}.
\eea
Here, $\Neighbors$ is the neighbor set for a mini-batch node, another random mini-batch of $n$ nodes sampled from $\cV^*$. Note that $|\Neighbors| \ll |\cV^*|=N$.

@@

\begin{algorithm}[t]
\caption{Sequential version of SG-MCMC for a-MMSB}\label{alg}
\begin{algorithmic}[1]
\STATE Initialize $\pi , \bt, \phi, \ta$
\WHILE {sampling}
\STATE Sample a mini-batch of vertex pairs, \Minibatch, from $E$
\FOR {each vertex in \Minibatch}
	\STATE Sample a mini-batch of vertices, $\Neighbors$, from $\Vertices$
    \STATE Update $\phi_{a}$ using Eqn~\ref{eqn:local_update}
  %  $\phi_{a}^* \law \left| \phi_{a} + \f{\ep_{\phi}}{2} \left( \al - \phi_{a} + \f{N}{|\Neighbors|} \sum_{b \in \Neighbors} G_{\phi_{a}}\right) + (\phi_{a})^{\ha} \xi_{a}\right|$
    \STATE Obtain $\pi_a$ from $\phi_a^*$ \ENDFOR
\FOR {$k = 1,\dots,K$}
    \STATE Update $\ta_{k}$ using Eqn~\ref{eqn:global_update}
   % $\ta_{k}^* \law \left| \ta_{k} + \f{\ep_\ta}{2} \left( \eta - \ta_{k} + h(\cE_t) \sum_{(a,b) \in \cE_t} G_{\ta_{k}}(a,b) \right) + (\ta_{k})^{\ha}\xi_{k} \right|$
    \STATE Obtain $\bt_k$ from $\ta_k^*$
\ENDFOR
\ENDWHILE
\end{algorithmic}
\label{algorithm}
\end{algorithm}

@@

\begin{comment}
Assortative mixed-membership stochastic blockmodel (a-MMSB) [1] is a special case of MMSB [8] that models the group-structure in a network of $N$ nodes. In particular, each node $a$ in the node set $\cV^*$ has a $K$-dimension probability distribution $\pi_a$ of participating in the $K$ members of the community set $\cK$. For every possible peer $b$ in the network, each node $a$ randomly draws a community $z_{ab}$. If a pair of nodes ($a,b$) in the edge set $\cE^*$ are in the same community: $z_{ab}=z_{ba} = k$, then they have a significant probability $\bt_k$ to connect, i.e., $y_{ab}=1$. Otherwise this probability is small. Each community has its connection strength $\beta_{k} \in (0,1)$ which explains how likely its members are linked to each other.

@@

\bea
\ta_{ki}^* \law \left| \ta_{ki} + \f{\ep}{2} \left\{ \eta - \ta_{ki} + h(\Minibatcht)\sum_{(a,b) \in \Minibatcht}g_{ab}(\ta_{ki})\right\} \right.\nn \\ \left.+ (\ta_{ki})^{\ha}\xi_{ki} \right|, \label{eqn:global_update}
\eea
here $\Minibatcht$ is a mini-batch of $n_t$ node pairs sampled from $\cE^*$ for which we use the following strategy.

\textbf{Stratified sampling:} considering that the number of links is much smaller than that of non-links, we can reduce the variance of the gradient using stratified sampling, similar to the method used in [1]. For this, at every iteration we first randomly select a node $a$ and then toss a coin with probability 0.5 to decide whether to sample link edges or non-link edges for node $a$. If it is a link, we assign all of the link edges of node $a$ to $\cE_{n_t}$. Otherwise, i.e. if it is non-link, we uniformly sample a mini-batch of $N/m$ non-link edges from the entire set of non-link edges and assign it to $\cE_{n_t}$. Here, the $m$ is a hyper-parameter. Note that the size of $|\cE_{n_t}|$ will thus be much smaller than the total number of $N(N-1)/2$ edges when $m$ is reasonably large. Then, to ensure that the gradient is unbiased, a \textit{scaling parameter} $h(\cE_{n_t})$ is multiplied. Specifically, $h(\cE_{n_t})$ is set to $N$ when $\cE_{n_t}$ is a set of link edges and to $mN$ otherwise.

@@

Then we obtain the normalization term by $Z_{ab}^{(y)} = \sm{k}{K}f_{ab}^{(y)}(k)$. 
Integrating out the expectation in Eqn. \eqref{eqn:grad_local}, we obtain 
\bea
g_{ab}(\phi_{ak}) = \f{f_{ab}^{(y)}(k)}{Z_{ab}^{(y)}\phi_{ak}} - \f{1}{\phi_a}.
\eea 
Plugging this to Eqn. \ref{eqn:sgrld_update}, we obtain the SGRLD update rule for the local parameter $\phi_{ak}$
\bea
\phi_{ak}^* \law \left| \phi_{ak} + \f{\ep}{2} \left( \al - \phi_{ak} + \f{N}{|\Neighbors|} \sum_{b \in \Neighbors} g_{ab}(\phi_{ak})\right) \right. \nn \\ \left. + (\phi_{ak})^{\ha} \xi_{ak}\right|.\label{eqn:local_update}
\eea
Here, the $\Neighbors$ is the neighbor set for a mini-batch node, another random mini-batch of $n$ nodes sampled from $\cV^*$. Note that $|\Neighbors| \ll |\cV^*|=N$.

>>>>

% In this section, we briefly review the Bayesian model we are considering in this paper.
\subsection{Assortative Mixed-membership Stochastic Blockmodels (a-MMSB)}
The assortative mixed membership stochastic blockmodel (a-MMSB)~\cite{gopalan2012scalable} is a special case of MMSB~\cite{airoldi2009mixed} that models the group structure in a network of $N$ vertices. Consider a set  $\cV$ containing all the vertices in the graph, and a set $\cE$ containing all the linked edges between pairs of vertices. Each vertex $a$ in the vertex set $\cV$ has a $K$-dimensional probability distribution $\pi_a$ of participating in the $K$ members of the community set $\cK$.  For every possible peer $b$ in the network, each vertex $a$ randomly draws a community $z_{ab}$. If a pair of vertices ($a,b$) in the edge set $\cE$ are in the same community, i.e., $z_{ab}=z_{ba} = k$, then they have a significant probability $\bt_k$ to connect, i.e., $y_{ab}=1$. Otherwise this probability is small. Each community has its connection strength $\beta_{k} \in (0,1)$ which reflects how likely its members are linked to each other. The whole generative process of a-MMSB is then described by:
\benum
\item For each community $k$, draw community strength $\bt_k \sim \mbox{Beta}(\eta)$
\item For each vertex $a$, draw community memberships $\pi_a \sim \mbox{Dirichlet}(\al)$
\item For each pair of vertices $a$ and $b$,
\benum
	\item Draw interaction indicator $z_{ab} \sim \pi_a$
    \item Draw interaction indicator $z_{ba} \sim \pi_b$
    \item Draw link $y_{ab} \sim \mbox{Bernoulli}(r)$, where $r = \bt_k$ if $z_{ab}=z_{ba}=k$, and $r=\dt$ otherwise.
\eenum
\eenum
$\eta$ and $\alpha$ are parameters to the Beta and Dirichlet distribution functions. $\dt$ (usually small) is a parameter for the algorithm.

\newcommand{\Vertices}{\ensuremath{\cV{}}\xspace}
\newcommand{\Edges}{\ensuremath{\cE{}}\xspace}
\newcommand{\Heldout}{\ensuremath{\cE_h{}}\xspace}
\newcommand{\SGLDMinibatch}{\ensuremath{\cD_n{}}\xspace}
\newcommand{\Minibatch}{\ensuremath{\cE_n{}}\xspace}
\newcommand{\Minibatcht}{\ensuremath{\cE_n{}}\xspace}
\newcommand{\Neighbors}{\ensuremath{\cV_n{}}\xspace}

\subsection{Stochastic Gradient Markov Chain Monte Carlo}
The algorithm we are working on in this paper is based on stochastic gradient Langevin dynamics (SGLD)~\cite{welling2011bayesian}. SGLD applies the following update rule to obtain samples from a posterior distribution $p(\ta|\cX) \propto p(\cX|\ta) p(\ta)$ of $N$ i.i.d. data points $\cX=\{x_i\}_{i=1}^{N}$:
\bea
\ta^* \law \ta + \f{\epsilon_t}{2}\left(\nabla_{\ta}\log p(\ta_t)+N\bar{g}(\ta;\SGLDMinibatch)\right) + \xi_t, \label{eqn:sgld_update}
\eea
where $\xi_t \sim \cN(0,\epsilon_t )$ with $\ep_t$ the step size, $\SGLDMinibatch$ a mini-batch of size $n$ sampled from $\cX$, and $\bar{g}(\ta;\SGLDMinibatch)$ the mean stochastic gradient, i.e., $\frac{1}{|\SGLDMinibatch|}\sum_{x\in \SGLDMinibatch}^{}\grad_{\ta} \log p(x|\ta)$. As the step size goes to zero by a schedule satisfying $\sm{t}{\infty}\ep_t = \infty$ and $\sm{t}{\infty}\ep_t^2 < \infty$, SGLD samples from the true posterior distribution. One benefit of using SGLD is that we do not need the Metropolis-Hastings (MH) accept-reject tests since the rejection probability goes to zero as the step size tends to zero. Although in practice we often do not reduce the stepsize to zero to obtain better mixing (thus resulting in some bias), we obtain good performance by drawing many more samples per unit time.

SGLD originated from the Langevin Monte Carlo (LMC)~\cite{girolami2011riemann}, where unlike SGLD the gradient is computed exactly by using all data points. Then, Metropolis-Hastings accept-reject tests are applied. Comparing to LMC, SGLD only requires to process a mini-batch $\SGLDMinibatch$ at each iteration and ignores the MH test, and thus the computation complexity substantially reduces from $\cO(N)$ to $\cO(n)$.

@@

\paragraph{\textbf{Sampling global parameters}} The update rule for global parameter $\theta$ is
\bea
\ta_{ki}^* \law \left| \ta_{ki} + \f{\ep}{2} \left( \eta - \ta_{ki} + h(\Minibatcht)\sum_{(a,b) \in \Minibatcht}g_{ab}(\ta_{ki})\right) \right.\nn \\ \left.+ (\ta_{ki})^{\ha}\xi_{ki} \right|, \label{eqn:global_update}
\eea
where the gradient in $\theta$
\bea
g_{ab}(\ta_{ki})
&=& \f{f_{ab}^{(y)}(k,k)}{Z_{ab}^{(y)}} \left(\f{|1-i-y|}{\ta_{ki}} - \f{1}{\sum_j\ta_{kj}} \right),
\eea
here $\Minibatcht$ is a mini-batch of $n_t$ vertex pairs sampled from $\Edges$; $h(\Minibatcht)$ is a weight factor to scale the effect of a mini-batch towards the full network; and 
\bea
f_{ab}^{(y)}(k,l)=
\begin{cases}
\bt_k^y(1-\bt_k)^{(1-y)}\pi_{ak}\pi_{bk}, & \mbox{if } k=l\\
\dt^{y}(1-\dt)^{(1-y)} \pi_{ak} \pi_{bl}, & \mbox{if } k \neq l \nn.
\end{cases}
\label{eqn:case}
\eea
$Z_{ab}^{(y)}$ is the normalization constant which we can compute in $\cO(K)$ time \cite{LiAW15}.

\paragraph{\textbf{Sampling local parameters}}
The update rule for the local parameters $\phi$ is
\bea
\phi_{ak}^* \law \left| \phi_{ak} + \f{\ep}{2} \left( \al - \phi_{ak} + \f{N}{|\Neighbors|} \sum_{b \in \Neighbors} g_{ab}(\phi_{ak})\right) \right. \nn \\ \left. + (\phi_{ak})^{\ha} \xi_{ak}\right|,
\label{eqn:local_update}
\eea
where the gradient in $\phi$
\bea
g_{ab}(\phi_{ak}) = \f{f_{ab}^{(y)}(k)}{Z_{ab}^{(y)}\phi_{ak}} - \f{1}{\sum_j\phi_{aj}}.
\eea
Here, $\Neighbors$ is the neighbor set for a mini-batch node, another random mini-batch of $n$ nodes sampled from $\cV$. Note that $|\Neighbors| \ll |\cV|=N$.

@@

\begin{algorithm}[t]
\caption{Sequential version of SG-MCMC for a-MMSB}\label{alg}
\begin{algorithmic}[1]
\STATE Initialize $\pi , \bt, \phi, \ta$
\WHILE {sampling}
\STATE Sample a mini-batch of vertex pairs, \Minibatch, from $\cE$
\FOR {each vertex in \Minibatch}
	\STATE Sample a mini-batch of vertices, $\Neighbors$, from $\Vertices$
    \STATE Update $\phi_{a}$ using Eqn~\ref{eqn:local_update}
  %  $\phi_{a}^* \law \left| \phi_{a} + \f{\ep_{\phi}}{2} \left( \al - \phi_{a} + \f{N}{|\Neighbors|} \sum_{b \in \Neighbors} G_{\phi_{a}}\right) + (\phi_{a})^{\ha} \xi_{a}\right|$
    \STATE Obtain $\pi_a$ from $\phi_a^*$ \ENDFOR
\FOR {$k = 1,\dots,K$}
    \STATE Update $\ta_{k}$ using Eqn~\ref{eqn:global_update}
   % $\ta_{k}^* \law \left| \ta_{k} + \f{\ep_\ta}{2} \left( \eta - \ta_{k} + h(\cE_t) \sum_{(a,b) \in \cE_t} G_{\ta_{k}}(a,b) \right) + (\ta_{k})^{\ha}\xi_{k} \right|$
    \STATE Obtain $\bt_k$ from $\ta_k^*$
\ENDFOR
\ENDWHILE
\end{algorithmic}
\label{algorithm}
\end{algorithm}

@@

\begin{comment}
Assortative mixed-membership stochastic blockmodel (a-MMSB) [1] is a special case of MMSB [8] that models the group-structure in a network of $N$ nodes. In particular, each node $a$ in the node set $\cV$ has a $K$-dimension probability distribution $\pi_a$ of participating in the $K$ members of the community set $\cK$. For every possible peer $b$ in the network, each node $a$ randomly draws a community $z_{ab}$. If a pair of nodes ($a,b$) in the edge set $\cE$ are in the same community: $z_{ab}=z_{ba} = k$, then they have a significant probability $\bt_k$ to connect, i.e., $y_{ab}=1$. Otherwise this probability is small. Each community has its connection strength $\beta_{k} \in (0,1)$ which explains how likely its members are linked to each other.

@@

\bea
\ta_{ki}^* \law \left| \ta_{ki} + \f{\ep}{2} \left\{ \eta - \ta_{ki} + h(\Minibatcht)\sum_{(a,b) \in \Minibatcht}g_{ab}(\ta_{ki})\right\} \right.\nn \\ \left.+ (\ta_{ki})^{\ha}\xi_{ki} \right|, \label{eqn:global_update}
\eea
here $\Minibatcht$ is a mini-batch of $n_t$ node pairs sampled from $\cE$ for which we use the following strategy.

\textbf{Stratified sampling:} considering that the number of links is much smaller than that of non-links, we can reduce the variance of the gradient using stratified sampling, similar to the method used in [1]. For this, at every iteration we first randomly select a node $a$ and then toss a coin with probability 0.5 to decide whether to sample link edges or non-link edges for node $a$. If it is a link, we assign all of the link edges of node $a$ to $\cE_n$. Otherwise, i.e. if it is non-link, we uniformly sample a mini-batch of $N/m$ non-link edges from the entire set of non-link edges and assign it to $\cE_n$. Here, the $m$ is a hyper-parameter. Note that the size of $|\cE_n|$ will thus be much smaller than the total number of $N(N-1)/2$ edges when $m$ is reasonably large. Then, to ensure that the gradient is unbiased, a \textit{scaling parameter} $h(\cE_n)$ is multiplied. Specifically, $h(\cE_n)$ is set to $N$ when $\cE_n$ is a set of link edges and to $mN$ otherwise.

@@

Then we obtain the normalization term by $Z_{ab}^{(y)} = \sm{k}{K}f_{ab}^{(y)}(k)$. 
Integrating out the expectation in Eqn. \eqref{eqn:grad_local}, we obtain 
\bea
g_{ab}(\phi_{ak}) = \f{f_{ab}^{(y)}(k)}{Z_{ab}^{(y)}\phi_{ak}} - \f{1}{\phi_a}.
\eea 
Plugging this to Eqn. \ref{eqn:sgrld_update}, we obtain the SGRLD update rule for the local parameter $\phi_{ak}$
\bea
\phi_{ak}^* \law \left| \phi_{ak} + \f{\ep}{2} \left( \al - \phi_{ak} + \f{N}{|\Neighbors|} \sum_{b \in \Neighbors} g_{ab}(\phi_{ak})\right) \right. \nn \\ \left. + (\phi_{ak})^{\ha} \xi_{ak}\right|.\label{eqn:local_update}
\eea
Here, the $\Neighbors$ is the neighbor set for a mini-batch node, another random mini-batch of $n$ nodes sampled from $\cV$. Note that $|\Neighbors| \ll |\cV|=N$.
